---
title: "Interest rate dispersion in commercial loans"
author: '"SA Moreno, AR Vargas, JS Vélez"'
date: "5 de septiembre de 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
library(knitr)
library(dplyr)
library(ggplot2)
library(kableExtra, quietly = TRUE)
suppressMessages(library(lmtest))
suppressMessages(library(sandwich))
suppressMessages(library(stargazer))
library(quantreg)
read_chunk('main.R')


```


## Introduction


## Literature


## Data

<!--
  


#### Propuesta á  la ACH (2013)
1. Documentar la importancia de la dispersión (magnitud, cambio en el tiempo, relación con estructura de mercado, con fusiones, etc.). Para poder hablar de dispersión tenemos que estar muy seguros de que nos quedamos con créditos homogéneos.
2. Identificar tipos de firmas que se benefician más de los descuentos de los bancos.
3. Testear que tanto de la dispersión se debe a discriminación de precios. 

--->


```{r data wrangling, include=F}
<<dataWrangling>>


```


## Empirical strategy


### Boxplot

The first thing we do is a boxplot that shows the cross-sectional dispersion of margins in commercial interest rates for every quarter in the data. We define margin as the difference between the interest rate negotiated at the start of the loan and the lender's opportunity cost. Below, we present the results obtained using the 5-year treasury bond rate as opportinity cost. However, rhe results are robust to several measures of opportunity cost (1 and 10-year treasury bond rate, the usury rate and the average certificate of deposit rate).The patterns of dispersion over time define three periods. 

A first period, spanning over 4 years, that stars with the first quarter of the data, that is, the last quarter of 2005, and ends with the second quarter of 2009. This period is characterized by high dispersion (large ranges) manifested, mostly, through exceptionally low margins that lie very far from the median.[^1] A second period, going from 2009Q3 to 2011Q2 in which there is relatively low dispersion, as margins are always smaller, in absolute value, than 15%. Finally, after 2011Q3 there is a surge in dispersion, this time caused by unusually large margins. Overall, the figure reveals a steady increase in the amount of dispersion caused by unsually large margins.

These distinct periods overlap with changes in policy that may, to some extent, explain the different patterns of dispersion that arise. For instance, starting 2007Q3 **ACÁ VA LO QUE ESTÁ ESCRIBIENDO RAFAEL SOBRE SARC.**

Similarly, in 2009 Congress passed Law 1328 which aim was to protect consumers of financial services (reducing search costs by increasing trasparency and information). The enactment of this law coincides with the period of lowest dispersion. **EXPANDIR**


Finally the third period coincides with a historic increase in the concentratio of the banking industry. **EXPANDIR Y CITAR FEDESARROLLO**


  




[^1]: We define range  as $Q3 +1.5\text{IQR} - [Q1-1.5\text{IQR}]$ where $Qi$ is a quartile and IQR is the inter-quartile range.


```{r boxplot, }

<<boxplot>>
  
```


### Variance decomposition

Next we want to study the importance of consumer-based pricing, that is, we want to quantify how much of the observed dispersion is due to the bank's ability to extract willingness to pay from the borrowers based on the borrower's observed characteristics. To do so we decompose the observed dispersion between four groups of covariates. In particular, we regress the margins $y_{i}$, defined as the negotiated rate minus the opportunity cost (5-yeard Treasury Bond Rate), on 4 groups of categorical variables $X$: quarter dummies, loan characteristics, borrower characteristics and market structure dummies. 



Specifically the equation we take to the data is 

$$y_{i}=\alpha + \sum_{x=1}^4\sum_{j=1}^{J_x}\mathbf{1}(x_i=j)\beta_{xj}+ e_i$$


Quarter dummies account for the importance of variation in the 5-year Treasury Bond rate dispersion. As loan characteristics, we use term, amount and a dummy that describes wether the loan was issued by the firm's the home-bank.[^2] We include as characteristics of the borrowing firms, revenue, debt, credit score, productivity and sector dummies.[^3] Finally, we include city dummies, a measure of market concentration and the size of the lender. The contiunous characterstics are discretized by percentiles 10th through 90th.

[^2]: The definition of home-bank about here.
[^3]: Brief reminder of what productivity means in this context. 

**Commenting the table**: **TO BE EXPANDED**

Table 1 shows the amount of variation explained by the dummies. Alternatively, table 2 shows the same information, but grouping the dummies in the 4 groups described above. 

```{r variance decomposition, warning=F}

<<varianceDecomposition>>
  
```


- At most the observed variables explain `r 100-round(min(tableanova[13,]),digits=2)*100`% of the variation which happens in `r colnames(tableanova[which(tableanova[13,]== min(tableanova[13,]))])`

- Reassuringly, across all periods the most important variables in explaining the dispersion of interest rates are term and loan size, which jointly explain up to `r round(max(tableanova[2,]+tableanova[3,]),digits=2)*100`%. More over, they have become more important in explaining the interest rates.

- The quarter dummies explain between `r round(min(tableanova[1,])*100,digits=2)`% and `r round(max(tableanova[1,]*100), digits=2)`%  of the dispersion. However, there is a clear fall in quarter-specific pricing between periods 1 and 3.

- A seemingly suprising fact is that the credit score has very little to do with dispersion. Moreover, at the end of the sample it explains a third of the explained variation in the begining. The reason behind this, is that the vast majority (`r round(sum(mainDf$calificacion=='A')/length(mainDf$calificacion)*100,digits=2)`%) of credits get the highest score.

- As expected the portion of dispersion explained by city dummies remains constant over time.

- The market structure explains much more of the price dispersion in the first two periods. Concentration measures explain very little of the dispersion (we tried 3 alternate measure of HHI: loans, #credits, #branches), however, the raw size of the lender, measured as number of branches nationwide, is capable of explaining between `r round(min(tableanova[12,])*100,digits=2)`% and `r round(max(tableanova[12,]*100), digits=2)`%  of the dispersion. It is worthwile mentioning that its importances halfed by the end of the sample.





### Who receives the discounts?


In this section we look at the relation between observed characteristics of loans, borrowers and markets, and the margins that the lenders can make. We estimate the following equation 


$$\text{Margin}_i=X_i\beta+\varepsilon_i$$

where $\text{Margin}_i$ is the difference between the negotiated rate and the 5-year Treasury bond rate and $X_i$ is the set of observed characteristic.

```{r who has negosh, results='asis', warning=FALSE}

<<negosh>>
  
```


Table 3 shows the result of regressing margins on observed loan, borrower, lender and market characteristics. Most of the estimated signs are as expected. More productive firms, on average, negotiate lower rates which has a negative impact on margins. Similarly, firms earning more revenue can negotiate higher discounts. Holding everything else constant, margins on loans to firms with 1% more revenue are 20 basis points smaller. Lenders make larger margins on indebted firms. Additional 1% debt at the moment of  negotiating the loan is correlated with 4 basis points larger margins. As expected, relative to the best credit score (A) all other scores imply a penalty for the borrowers, so lenders earn larger margins with loans rates B, C, D or E. 

Regarding market characteristics, we find at first a positive correlation between concentration (HHI) and margin, but once we include city fixed effects the sign of the estimate changes. This means that lenders in cities with more concentrated markets, but otherwise similar, actually make smaller margins. In contrast, we find a negative correlation between the absolute size of the bank and the margin.

Longer terms and larger amounts are correlated with smaller margins. After including all fixed effects, an additional month correlates with a decrease in margin of 2 basis points. Similarly, a 1% larger amount implies a 1.70 percent points smaller margin. 

An interesting result has to do with the estimate on the "home bank" dummy. The positive sign means lenders make larger margins on recurrent customers. This premium charged to their clients is on average 37 basis points. This suggest that the banks use their bargaining power to extract willingness to pay from borrowers. As an informal check, ee run the same regression with a subsample of same-day loans, which arguably leave the borrower with very little power to neogtiate better rates, and find that the estimates on home bank are 5 times larger (1.84) and statistically significant.[^4] 

[^4]: Regression in the appendix.**NOT YET**




3.  **Quantile regression** of transaction rate (or some kind of measure of profit margin) on market, firm and bank characteristics. Whatever's left is interpreted as price discrimination. 


```{r quantile reg, warning=FALSE}

<<quantreg>>
  
```



4. **Test**: Marginal effects of observable characteristics on rates should be close to zero for low rates (high negosh) than for medium rates (medium negosh).



